\MakeUppercase{Linear versus Centered Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2021-09-27 lin-vs-cen.tex
Getting somewhere

\usepackage[mathlines]{lineno}
\setlength{\linenumbersep}{2em}
\linenumbers
% \rightlinenumbers
\linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

\DeclareMathOperator{\ddiv}{div}
\DeclareMathOperator{\hist}{p}

@@

@@

The $k\times k$ grid graph $G_k$ is the graph with vertex set $V(G_k):=\{1,\ldots,k\}^2$ and that contain an edge with endpoints $(v,w)$ and $x,y$ if and only $|v-x|+|w-y|=1$.  A graph $H$ is a $k\times k$ \emph{pseudogrid} if there exists a set $\mathcal{P}$ of vertex-disjoint paths in $G$ such that the quotient graph $H':=H/\mathcal{P}$ is isomorphic to a subdivision of $G_k$.  Since $G_k$ has maximum degree $4$ any graph $G$ that contains a $G_k$-minor contains a subgraph $H$ that is a $k\times k$ pseudogrid.

@@

\begin{proof}
  Let $\varphi:V(H)\to\{1,\ldots,c\}$ be an arbitrarily $c$-colouring of some $k\times k$ pseudogrid $H$ for some $c\le \epsilon k$, where $\epsilon$ is a small positive constant.  We will show that $H$ contains a path $P$ which has no center with respect to $c$.  First we introduce some notations:

@@

  For each colour $\alpha\in\{1,\ldots,c\}$, define the \emph{multiplicity} of $\alpha$ as the number of vertices and edges of $G_k$ that include $\alpha$ in their colour set.  Fix some large integer constant $A$ and say that $\alpha$ is \emph{infrequent} if it has multiplicity at most $A$.

  If there exists some infrequent colour $\alpha\in\{1,\ldots,c\}$, we identify every row and column of $G_k$ that contains an edge or vertex with $\alpha$ in its colour set.  We then \emph{remove} each of these (at most $A$) rows and columns from $H$ by removing all edges of $H$ in $P_e$ for each $e$ in the relevant row (column) of $G_k$ and then discarding any isolated vertices.  After doing this for each row and column containing $\alpha$, the graph $H$ may be disconnected.  We discard any isolated vertices and any path $P_{v}$ that is disconnected from the rest of $H$.

  The resulting graph $\tilde{H}$ contains a $k'\times k'$ pseudogrid for some $k' \le k-A$ and contains no vertex of colour $\alpha$.  We then set $H\gets \tilde{H}$, $k\gets k'$ and repeat this process until eventually obtain a new $k'\times k'$ pseudogrid $\tilde{H}$ for some $k'\ge k-cA\ge (1-\epsilon A)k \ge k/2$, for $\epsilon \le 1/2A$.  From this point onward that that there is no infrequent colour, i.e., every colour $\alpha\in\{1,\ldots,c\}$ has multiplicity at least $A$.

  Next we run a two stage process that, for each $\alpha\in\{1,\ldots,c\}$, will identify a pair of vertices $v_\alpha,w_\alpha$ in $H$ whose colour is $\alpha$.  These pairs are carefully chosen so that we can construct a simple path $P$ in $H$ that contains $\bigcup_{\alpha\in\{1,\ldots,c\}} \{v_\alpha,w_\alpha\}$.

  \begin{compactenum}[Stage 1]
    \item We begin this stage by defining each edge and vertex of $G_k$ as being initially \emph{unmarked}.  Fix some constant $B>0$.  We iterate over each $\alpha\in\{1,\ldots,c\}$ in turn.  If $G_k$ contain a pair $(x_\alpha,y_\alpha)$ where each of $x_\alpha$ and $y_\alpha$ is an unmarked vertex or unmarked edge of $G_k$ with $\alpha$ in its colour set and the distance in $G_k$ between $x_\alpha$ and $y_\alpha$ is at least $B$ then Stage~1 \emph{succeeds} for $\alpha$.  In this case, we take $v_\alpha$ to be any vertex in $P_{x_\alpha}$ of colour $\alpha$ and $w_\alpha$ to be any vertex in $P_{y_\alpha}$ of colour $\alpha$.  We then \emph{mark} every vertex and edge of $G_k$ whose distance from $x_\alpha$ or $y_\alpha$ is less than $B$.

    If $G_k$ contains no such pair $(x_\alpha,y_\alpha)$, then we leave $v_\alpha$ and $w_\alpha$ undefined and say that Stage~1 \emph{failed} for colour $\alpha$.

    \item Let $S\subset\{1,\ldots,c\}$ be the set of colours $\alpha$ for which Stage~1 succeeded and let $\overline{S}:=\{1,\ldots,c\}\setminus S$ be the set of colours for which Stage~1 failed.  We construct a bipartite graph $K$ with parts $L(K):=\overline{S}$ and $R(K):=\bigcup_{\alpha\in S}\{x_\alpha,y_\alpha\}$.   For each marked vertex or edge $z$ of $G_k$, let $s(z)$ be the vertex in $R(k)$ that is closest to $z$ in $G_k$.\todo{Gotta be really precise here to avoid degree 3 issue.}  The edge set of $K$ contains an edge $\alpha\beta$ precisely when there is a marked edge or vertex $z$ of $G_k$ such that $\alpha\in\phi(z)$ and $s(z)=\beta$.

    \begin{clm}
      $K$ contains two matchings $M_1$ and $M_2$ such that each vertex in $L(K)$ appears in each of $M_1$ and $M_2$ and each vertex in $R(k)$ appears in at most one of $M_1$ or $M_2$.
    \end{clm}

    \begin{proof}
      Use Hall's Mathcing Theorem/Lemma.
    \end{proof}

    Finally, for each $\alpha\in\overline{S}$, we define $x_\alpha$ to be the the unique element of $R(K)$ such that $M_1$ includes the edge $\alphax_\alpha$ and we define $y_\alpha$ to be the unique element of $R(K)$ such that $M_2$ includes the edge $\alpha y_\alpha$.  We define $v_\alpha$ and $w_\alpha$ as before.

    Now, let $R:=\bigcup_{\alpha\in\{1,\ldots,c\}} \{x_\alpha,y_\alpha\}$.

@@

    \begin{clm}
        Each $B/2\times B_/2$ subgrid of $G_k$ contains at most four elements of $R$.
    \end{clm}

@@

    \begin{lem}
        $G_k$ contain a simple path $P$ that contains every edge and vertex in $R$.  Furthermore, $P$ bends at each bent vertex in $R$ and $P$ goes straight at each straight vertex in $R$. 
    \end{lem}

\end{proof}

@@

Two words $s$ and $t$ are \emph{anagrams} of each other if $s$ is a permutation of $t$.  A single word $w:=w_1,\ldots,w_{2r}$ is \emph{anagramish} if its first half $w_1,\ldots,w_r$ and its second half $w_{r+1},\ldots,w_{2r}$ are anagrams of each other.  A word is \emph{anagram-free} if it does not contain an anagramish factor.  Anagramish words are also known as \emph{abelian squares} and anagram-free words are also known as \emph{abelian square free} words \cite{keranen:abelian,keranen:powerful}, \emph{strongly nonrepetitive} words \cite{pleasants:non-repetitive}, or \emph{strongly asymmetric} sequences \cite{evdokimov:strongly,evdokimov:strongly2}.

In 1961, \citet{erdos:some} asked if there exist arbitrarily long words over an alphabet of size $4$ that are anagram-free.\footnote{This was an incredibly prescient question since it is not at all obvious that there exist arbitrarily long anagram-free words over \emph{any} finite alphabet. The only justification for choosing the constant $4$ is that a short case analysis rules out the possibility of length-$8$ anagram-free words over an alphabet of size $3$.}  In 1968 \citet{evdokimov:strongly,evdokimov:strongly2} showed the existence of arbitrarily long anagram-free words over an alphabet of size $25$ and in 1971 \citet{pleasants:non-repetitive} showed an alphabet of size $5$ is sufficient.  Erd\H{o}s's question was not fully resolved until 1992, when \citet{keranen:abelian} answered it in the affirmative.

A path $v_1,\ldots,v_{2r}$ in a graph $G$ is \emph{anagramish} under a vertex $c$-colouring $\phi:V(G)\to\{1,\ldots,c\}$ if $\phi(v_1),\ldots,\phi(v_{2r})$ is an anagramish word.  The colouring $\phi$ is an \emph{anagram-free colouring} of $G$ if no path in $G$ is anagramish under $\phi$.  The minimum integer $c$ for which $G$ has an anagram-free vertex $c$-colouring is called the \emph{anagram-free chromatic number} of $G$ and is denoted by $\afcn(G)$.

For a graph family $\mathcal{G}$, $\afcn(\mathcal{G}):=\max\{\afcn(G):G\in\mathcal{G}\}$ or $\afcn(\mathcal{G}):=\infty$ if the maximum is undefined. The results on anagram-free words discussed in the preceding paragraph can be interpreted in terms of $\afcn(\mathcal{P})$ where $\mathcal{P}$ is the family of all paths. Indeed, Ker\"anen's result shows that $\afcn(\mathcal{P})\le 4$. Slightly more complicated than paths are trees.  \citet{wilson.wood:anagram-free} showed that $\afcn(\mathcal{T})=\infty$ for the family $\mathcal{T}$ of trees and \citet{kamcev.luczak.ea:anagram-free} showed that $\afcn(\mathcal{T}_2)=\infty$ even for the family $\mathcal{T}_2$ of binary trees.

One positive result in this context is that of
\citet{wilson.wood:anagram-free}, who showed that every tree $T$ of pathwidth $p$ has $\afcn(T)\le 4p+1$. Since trees are graphs of treewidth $1$ it is natural to ask if this result can be extended to show that every graph $G$ of treewidth $t$ and pathwidth $p$ has $\afcn(G)\le f(t,p)$ for some function $f:\N^2\to\N$.  \citet{carmi.dujmovic.ea:anagram-free} showed that such a generalization is not possible for any $t\ge 3$ by giving examples of $n$-vertex graphs of pathwidth $3$ (and treewidth $3$) with $\afcn(G)\in\Omega(\log n)$.  The obvious remaining gap left by these two works is graphs of treewidth $2$. Our main result is to show that $G_n$, the $2\times n$ square grid has $\afcn(G_n)\in\omega_n(1)$. Since $G_n$ has pathwidth 2, we have:

\begin{thm}\label{main_vertex}
    For every $c\in\N$, there exists a graph of pathwidth $2$ that has no anagram-free vertex $c$-colouring.
\end{thm}

\citet[Section 7.1]{wilson:anagram-free} conjectured that $\afcn(G_n)\in\omega_n(1)$, so this work confirms this conjecture.  Prior to the current work, it was not even known if the family of $n\times n$ square grids had anagram-free colourings using a bounded number of colours.

In a larger context, this lower bound gives more evidence that, except for a few special cases (paths \cite{evdokimov:strongly,pleasants:non-repetitive,keranen:abelian}, trees of bounded pathwidth \cite{wilson.wood:anagram-free}, and highly subdivided graphs \cite{wilson.wood:anagram-free2}), the qualitative behaviour of anagram-free chromatic number is not much different than that of treedepth/centered colouring \cite{nesetril.ossona:tree-depth}.  Very roughly: For most graph classes, every graph in the class has an anagram-free colouring using a bounded number of colors precisely when every graph in the class has a colouring using a bounded number of colours in which every path contains a colour that appears only once in the path.

% \todo[inline]{Statement of vertex colouring result (hopefully)}
%
% An \emph{edge $c$-colouring} $\varphi:E(G)\to\{1,\ldots,c\}$ is \emph{anagram-free} if $G$ has no path $v_0,\ldots,v_{2r}$ such that $\varphi(v_0v_1),\ldots,\varphi(v_{2r-1}v_{2r})$ is anagramish.  We also give an edge colouring version of \cref{main_vertex}.
%
% \begin{thm}\label{main_edge}
%     For every $c\in\N$, there exists a graph of $G$ of pathwidth $2$ that has no anagram-free edge $c$-colouring.
% \end{thm}

% \todo[inline]{Relation to Existing Work}

The remainder of this paper is organized as follows: \Cref{near_anagram_statement} gives some definitions and states a key lemma which shows, under a certain periodicity condition, that every sufficiently long word contains a factor that is $\epsilon$-close to being anagramish.
% In \cref{edge_colourings} we prove \cref{main_edge}.
In \cref{vertex_colourings} we use this key lemma to prove \cref{main_vertex}.  In \cref{near_anagram_proof} we prove the key lemma. \cref{reflections} concludes with some final remarks about the (non-)constructiveness of our proof technique.

\section{Periodicity in words}
\label{near_anagram_statement}

An \emph{alphabet} $\Sigma$ is a finite non-empty set and each element of $\Sigma$ is called a \emph{letter}. A \emph{word} over $\Sigma$ is a (possibly empty) sequence $w:=w_1,\ldots,w_n$ with $w_i\in\Sigma$ for each $i\in\{1,\ldots,n\}$. The \emph{length} $|w|$ of $w$ is the length, $n$, of the sequence.  A word of length at least $1$ is \emph{non-empty}.  The unique word of length $0$ is denoted by $\varepsilon$. For a word $w:=w_1,\ldots,w_n$, we let $\Sigma_w:=\{w_i:i\in\{1,\ldots,n\}\}$.  For each $1 \le i \le j\le n+1$, $w_i,w_{i+1},\ldots,w_{j-1}$ is called a \emph{factor} of $w$.\footnote{Note that this definition includes the empty factor $\varepsilon$ obtained when $i=j$.}  The set of all factors of $w$ is denoted by $F(w)$.

A \emph{language} $L$ over $\Sigma$ is a set of words over $\Sigma$.  A language $L$ is \emph{infinite} if $|L|=\infty$.  We let $\Sigma_L:=\bigcup_{w\in L}\Sigma_w$ and $F(L):=\bigcup_{w\in L}F(w)$.  The language $L$ is \emph{factor-closed} if $F(w)\subseteq L$ for each $w\in L$.  For any alphabet $\Sigma$ and any $k\in\N$, $\Sigma^k$ (the $k$-fold cartesian product of $\Sigma$ with itself) is the language consisting of all length-$k$ words over $\Sigma$.  The \emph{Kleene closure} $\Sigma^*:=\bigcup_{k=0}^\infty \Sigma^k$ is the language consisting of all words over $\Sigma$.

 % (Note the convention that $w[i\mathbin{:}j]$ does not include $w_j$, so $w[i\mathbin{:}j]$ has length $j-i$.)\todo{Is this notation still needed after revisions?}

Let $w:=w_1,\ldots,w_n$ be a word over an alphabet $\Sigma$ and, for each $a\in \Sigma$, define $|w|_a:=|\{i\in\{1,\ldots,n\}:w_i=a\}|$.  The \emph{Parikh vector} of $w$ is the integer-valued $|\Sigma|$-vector $\hist(w):=(|w|_{a}:a\in\Sigma)$ that is indexed by elements of $\Sigma$.  Observe that a word $w_1,\ldots,w_{2r}$ is anagramish if and only if $\hist(w_1,\ldots,w_r)=\hist(w_{r+1},\ldots,w_{2r})$ or, equivalently, $\hist(w_1,\ldots,w_r)-\hist(w_{r+1},\ldots,w_{2r})=\boldsymbol{0}$.
% Let $\|\cdot\|_1$ denote the L1 norm of a vector.\footnote{The L1 norm of a real-valued $d$-vector $\boldsymbol{v}:=(v_1,\ldots,v_d)$, $\|\boldsymbol{v}\|_1=\sum_{i=1}^d|v_i|$.}
For each $a\in\Sigma$, let $\delta_a(w):=|w_1,\ldots,w_r|_{a}-|w_{r+1},\ldots,w_{2r}|_{a}$ and let $\tau_a(w):=|\delta_a|$.
Then $\tau(w):=\sum_{a\in\Sigma}\tau_a(w)$ is a useful measure of how far a word is from being anagramish and that $\tau(w)=0$ if and only if $w$ is anagramish.

A word $w:=w_1,\ldots,w_n$ is \emph{$\ell$-periodic} over an alphabet $\Sigma\supseteq\Sigma_w$ if each length-$\ell$ factor of $w$ contains every letter in $\Sigma$.  A language $L$ is \emph{$\ell$-periodic}  if each of its words is $\ell$-periodic over $\Sigma_L$.  A language $L$ is \emph{periodic} if it is $\ell$-periodic for some finite $\ell$.  We make use of the following key lemma, which states that every sufficiently long $\ell$-periodic word contains arbitrarily long factors that are $\epsilon$-close to being anagramish.

\begin{lem}\label{near_anagram_fourier}
    For each $r_0,\ell\in\N$ and each $\epsilon>0$, there exists a positive integer $n$ such that each $\ell$-periodic word $w_1,\ldots,w_n$ contains a factor $w:=w_{i+1},\ldots,w_{i+2r}$ of length $2r \ge 2r_0$ such that $\tau(w)\le \epsilon r$.
\end{lem}

The proof of \cref{near_anagram_fourier} is deferred to \cref{near_anagram_proof}.  We now give some intuition as to how it is used.  The process of checking if a word is anagramish can be viewed as finding common letters in the first and second halves and crossing them both out.  If this results in a complete cancellation of all letters, then the word is an anagram.  \cref{near_anagram_fourier} tells us that we can always find a long factor $w$ where, after exhaustive cancellation, only an $\epsilon$-fraction of the original letters remain.
% Informally, the factor $w$ is $\epsilon$-close to being anagramish.

\cref{near_anagram_fourier} says that, if up to $\epsilon r$ letters in each half of $w$ were each allowed to cancel two letters of the same letters in the other half of $w$, then it would be possible to complete the cancellation process.  To achieve this type of one-versus-two cancellation in our setting, we decompose our coloured pathwidth-$2$ graph into pieces of constant size.  The vertices in each piece can be covered with one path or partitioned into two paths (see \cref{paths}).  In this way an occurrence $H_z$ of a particular coloured piece in one half can be matched with two like-coloured pieces $H_x$ and $H_y$ in the other half. We construct a single path $P$ that contains all vertices in $H_z$ and only half the vertices in each of $H_x$ and $H_y$.  In this way, the colours of vertices $P\cap H_z$ can cancel the colours of the vertices in $P\cap(H_x\cup H_y)$.

Since \cref{near_anagram_fourier} requires that the word $w$ be $\ell$-periodic, the following lemma will be helpful in obtaining words that can be used with \cref{near_anagram_fourier}. (Although \cref{periodicity} follows immediately from considerably stronger results in combinatorics on words---for example \citet[Proposition~1.5.12]{lothaire:algebraic}---we provide a proof here for the sake of completeness.)

\begin{lem}\label{periodicity}
    Let $L$ be an infinite factor-closed language over an alphabet $\Sigma$.  Then there exists an infinite periodic factor-closed language $M\subseteq L$.
\end{lem}

\begin{proof}
    Let $\Xi\subseteq\Sigma$ be any minimal subset of $\Sigma$ with the property that $M:=\Xi^* \cap L$ is infinite.  Observe that, since $L$ is factor closed, $M$ is also factor-closed.  Therefore $M$ is infinite and factor-closed.  All that remains is to show that $M$ is $\ell$-periodic for some integer $\ell$.

    Since $\Xi$ is minimal, $\Pi^*\cap L$ is finite for any strict subset $\Pi\subsetneq \Xi$.  Since $L$ is factor-closed, $\Pi^*\cap L$ is factor-closed.  Since $\Pi^*\cap L$ is factor-closed and finite, $|w|$ is finite for each $w\in\Pi^*\cap L$.
    Therefore, the value
    \begin{equation}
        \ell:=1+\max\{|w|: w\in \Pi^*\cap L,\, \Pi\subsetneq\Xi\} \label{l-def}
    \end{equation}
    is finite.  Now, consider any factor $f$ of any word in $M$ that does not contain every letter of $\Xi$, i.e., $f\in\Pi^*$ for some  $\Pi\subsetneq\Xi$.  Since $\Pi^*\cap L$ is factor closed, $f\in \Pi^*\cap L$ so, by \cref{l-def}, $|f| < \ell$.  Contrapositively, any factor of length at least $\ell$ of any word in $M$ contains every element of $\Xi$, so $M$ is $\ell$-periodic.
\end{proof}

\section{Proof of \cref{main_vertex}}
\label{vertex_colourings}

For each $n\in\N$, let $G_n$ be the $2\times n$ square grid with top row $a_0,\ldots,a_{n-1}$ and bottom row $b_0,\ldots,b_{n-1}$ (see \cref{g_n}).  For the sake of contradiction, suppose that there exists an integer $c$ such that $\afcn(G_n)\le c$ for each $n\in\N$.  Observe that there is a bijection between vertex $c$-colourings of $G_n$ and words of length $2n$ over the alphabet $\Sigma_0:=\{1,\ldots,c\}$ where a word $w:=w_0,\ldots,w_{2n-1}$ corresponds to the colouring $\phi_w:V(G_n)\to\{1,\ldots,c\}$ defined by
$\phi_w(a_i):=w_{2i}$ and $\phi_w(b_i):=w_{2i+1}$, for each $i\in\{0,\ldots,n-1\}$.

\begin{figure}
    \centering{
        \includegraphics{figs/g_n}
    }
    \caption{The graph $G_n$}
    \label{g_n}
\end{figure}

>>>>

\usepackage[mathlines]{lineno}
\setlength{\linenumbersep}{2em}
% \linenumbers
% \rightlinenumbers
% \linenumbers
\newcommand*\patchAmsMathEnvironmentForLineno[1]{%
 \expandafter\let\csname old#1\expandafter\endcsname\csname #1\endcsname
 \expandafter\let\csname oldend#1\expandafter\endcsname\csname end#1\endcsname
 \renewenvironment{#1}%
    {\linenomath\csname old#1\endcsname}%
    {\csname oldend#1\endcsname\endlinenomath}}%
\newcommand*\patchBothAmsMathEnvironmentsForLineno[1]{%
 \patchAmsMathEnvironmentForLineno{#1}%
 \patchAmsMathEnvironmentForLineno{#1*}}%
\AtBeginDocument{%
\patchBothAmsMathEnvironmentsForLineno{equation}%
\patchBothAmsMathEnvironmentsForLineno{align}%
\patchBothAmsMathEnvironmentsForLineno{flalign}%
\patchBothAmsMathEnvironmentsForLineno{alignat}%
\patchBothAmsMathEnvironmentsForLineno{gather}%
\patchBothAmsMathEnvironmentsForLineno{multline}%
}

@@

% \DeclareMathOperator{\ddiv}{div}
% \DeclareMathOperator{\hist}{p}
\DeclareMathOperator{\dist}{dist}

@@

\section{Preliminaries}

For a graph $G$ and a vertex $v\in V(G)$, $\deg_G(v):=|\{vw\in E(G)\}|$ denotes the degree of $v$ in $G$.  For a any $S\subseteq V(G)$, $\deg_G(S):=\sum_{v\in S}\deg_G(v)$ is the total degree of $S$.\todo{Surely this next lemma is well-known?}

\begin{lem}\label{hall_vees}
  Let $G$ be a bipartite graph with vertex parts $A$ and $B$ and having the property that, for any $A_0\subseteq A$, $|N_G(W)|\ge 2|A_0|$.  Then $G$ contains a subgraph $\Lambda$ in which each vertex of $A$ has degree exactly two and each vertex of $B$ has degree at most one.
\end{lem}

\begin{proof}
  Let $\Lambda$ be a subgraph of $G$ in which each vertex of $A$ has degree at most two, each vertex of $B$ has degree at most one, and that maximizes $\deg_\Lambda(A)$.  If $\deg_{\Lambda}(A) = 2|\Lambda|$ then there is nothing to prove. Assume therefore, for the sake of contradiction,  that there exists $v_0\in A$ with $\deg_\Lambda(v_0) < 2$.

  We say that a path $v_0,\ldots,v_m$ in $G$ is \emph{$\Lambda$-aternating} if $v_iv_{i+1}\in E(\Lambda)$ for all odd $i\in\{0,\ldots,m-1\}$ and $v_iv_{i+1}\not\in E(\Lambda)$ for all even $i\in\{0,\ldots,m\}$.  Observe that any such path has even length so that it ends at a vertex $v_m\in A$.  Otherwise, replacing the edges $\lfloor m/2\rfloor$ edges $\{v_1v_2,v_3v_4,\ldots,v_{m-2}v_{m-1}\}\in E(\Lambda)$ with the $\lceil m/2\rceil$ edges $\{v_0v_1,v_2v_3,\ldots,v_{m-1}v_m\}$ does not change $\deg_\Lambda(v)$ for any $v\in A_0\setminus\{v_0\}$ but does increase  $\deg_\Lambda(v_0)$, contradicting the assumption that $\Lambda$ maximizes $\deg_\Lambda(A)$.

  Let $A_0\subseteq A$ and $B_0\subseteq B$ be the subsets of $A$ and $B$, respectively, that can be reached from $v_0$ by $\Lambda$-alternating paths.
  Observe that, for any $x\in B_0$,  there is exactly one edge $vx$ of $\Lambda$ incident on $x$ and the other endpoint $v$ of this edge is in $A_0$.  Let $a$ be the number of edges of $\Lambda$ having one endpoint $A_0$ and one endpoint in $B_0$, so that $a=\deg_\Lambda(B_0)=|B_0|$.  Let $b$ be the number of edges of $\Lambda$ having one endpoint in $A_0$ and one endpoint in $B\setminus B_0$, so that $\deg_\Lambda(A_0)=a+b = |B_0|+b$.    Since $\deg_{\Lambda}(v) \le 2$ for each $v\in A_0$ and $\deg_\Lambda(v_0)<2$, $\deg_\Lambda(A_0)<2|A_0|$.  This yields the desired contradiction, since $|N_\Lambda(A_0)| \le \deg_\Lambda(A_0) = |B_0|+b = \deg_\Lambda(A_0)< 2|A_0|$.
\end{proof}

@@

The $k\times k$ grid graph $G_k$ is the graph with vertex set $V(G_k):=\{1,\ldots,k\}^2$ and that contain an edge with endpoints $(v,w)$ and $(x,y)$ if and only $|v-x|+|w-y|=1$.  A graph $H$ is a $k\times k$ \emph{pseudogrid} if there exists a set $\mathcal{P}$ of vertex-disjoint paths in $G$ such that the quotient graph $H':=H/\mathcal{P}$ is isomorphic to a subdivision of $G_k$.  Since $G_k$ has maximum degree $4$ any graph $G$ that contains a $G_k$-minor contains a subgraph $H$ that is a $k\times k$ pseudogrid.

@@

\begin{proof}
  Let $\varphi:V(H)\to\{1,\ldots,C\}$ be an arbitrarily $C$-colouring of some $k\times k$ pseudogrid $H$ for some $C\le \epsilon k$, where $\epsilon$ is a small positive constant.  We will show that $H$ contains a path $P$ which has no center with respect to $c$.  First we introduce some notations:

@@

  For each colour $\alpha\in\{1,\ldots,C\}$, define the \emph{multiplicity} of $\alpha$ as the number of vertices and edges of $G_k$ that include $\alpha$ in their colour set.  Fix some large integer constant $\alpha$ and say that $\alpha$ is \emph{infrequent} if it has multiplicity at most $\alpha$ and \emph{frequent} otherwise.

  \begin{clm}\label{all_frequent}
    $H$ contains a $k'\times k'$ pseudogrid in which every colour is frequent, for some $k'\ge (1-\epsilon \alpha)k$.
  \end{clm}

  \begin{proof}
    If there exists some infrequent colour $c\in\{1,\ldots,C\}$, we identify every row and column of $G_k$ that contains an edge or vertex with $c$ in its colour set.  We then \emph{remove} each of these (at most $A$) rows and columns from $H$ by removing all edges of $H$ in $P_e$ for each $e$ in the relevant row (column) of $G_k$ and then discarding any isolated vertices.  After doing this for each row and column containing $c$, the graph $H$ may be disconnected.  We discard any isolated vertices and any path $P_{v}$ that is disconnected from the rest of $H$.

    The remaining graph contains a $k'\times k'$ pseudogrid for some $k' \ge k-A$ and contains no vertex of colour $c$.  We then repeat this (at most $c$ times) until the resulting graph has no infrequent colour.  This leaves a graph that contains a $k'\times k'$ pseudogrid for some $k'\ge k-C\alpha\ge(1-\epsilon \alpha)k$ for which every colour is frequent.
  \end{proof}

  From this point onward, we assume that there is no infrequent colour, i.e., each colour $c\in\{1,\ldots,C\}$ has multiplicity at least $\alpha$.  Next we run a two stage process that, for each $c\in\{1,\ldots,C\}$, will identify a pair of vertices $v_c,w_c$ in $H$ whose colour is $c$.  These pairs are carefully chosen so that we will be able to construct a simple path $P$ in $H$ that contains $\bigcup_{c\in\{1,\ldots,C\}} \{v_c,w_c\}$.

  \begin{compactenum}[{Stage} 1.]
    \item We begin this stage by defining each vertex and edge of $G_k$ as being initially \emph{unmarked}.  Fix some integer constant $\beta>0$.  We iterate over each $c\in\{1,\ldots,C\}$ in turn.  If $G_k$ contain a pair $(x_c,y_c)$ where each of $x_c$ and $y_c$ is an unmarked vertex or unmarked edge of $G_k$ with $c$ in its colour set and the distance in $G_k$ between $x_c$ and $y_c$ is at least $B$ then Stage~1 \emph{succeeds} for $c$.  In this case, we take $v_c$ to be any vertex in $P_{x_c}$ of colour $c$ and $w_c$ to be any vertex in $P_{y_c}$ of colour $c$.  We then \emph{mark} every vertex and edge $z$ of $G_k$ such that $\dist_{G_k}(z, \{x,y\})\le \beta$.

    If $G_k$ contains no such pair $(x_c,y_c)$, then we leave $v_c$ and $w_c$ undefined and say that Stage~1 \emph{fails} for colour $c$.  Note that Stage~1 can only fail for $c$ if $G_k$ contains few than $B^2$ unmarked vertices/edges with $c$ in their colour sets.

    \item Let $S\subset\{1,\ldots,C\}$ be the set of colours for which Stage~1 succeeded and let $\overline{S}:=\{1,\ldots,C\}\setminus S$ be the set of colours for which Stage~1 failed.  We construct a bipartite graph $I$ with vertex parts $A:=\overline{S}$ and $B:=\bigcup_{c\in S}\{x_c,y_c\}$.   For each marked vertex or edge $p$ of $G_k$, let $s(p)$ be a vertex $q\in B$ that minimizes $\dist_{G_k}(p,q)$.\todo{Gotta be really precise here to avoid degree 3 issue.}  The edge set of $I$ contains an edge $xy$ precisely when there is a marked edge or vertex $z$ of $G_k$ such that $x\in\phi(z)$ and $s(z)=y$.

    First observe that, for any node in $y\in B$, the number of $x\in A$ such that $\dist_{G_k}(x,y) \le \beta$ is at most $8\beta^2$.   Therefore, the number of marked vertices/edges $x$ of $G_k$ with $s(x)=y$ is at most $8\beta^2$.  Since each colour $c\in A$ is frequent, this implies that $\deg_I(c)\ge \alpha/8\beta^2$.  Similarly, for each $y\in B$,  $\deg_{I}(y) \le 8\beta^12$.  This implies that, for each $A_0\subseteq A$, $|N_I(A_0)|\ge \alpha/64\beta^4$.   In particular, for any $\alpha$ and $\beta$ with $\alpha > 128\beta^4$,  $|N_I(A_0)|\ge 2|A_0|$.  Therefore, by \cref{hall_vees}, there exists a subgraph $\Lambda$ of $I$ in which each vertex of $A$ has degree $2$ and each vertex of $B$ has degree at most one.  Therefore, each $c\in A=\overline{S}$ has two $\Lambda$-neighbours $p$ and $q$ in $B$.  The edge $cp$ is present in $I$ because $G_k$ contains a vertex $x_c$ with $s(x_c)=p$ and $c\in \phi(x_c)$.  Similarly, the edge $cq$ is present in $I$ because $G_k$ contains a vertex $y_c$ with $s(y_c)=q$ and $c\in\phi(y_c)$.  This defines the pairs $\{x_c,y_c\}$ for each element in $\overline{S}$.
  \end{compactenum}

  Summarizing our work thus far, we have identified a subset  $R:=\bigcup_{c=1}^C\{x_c,y_c\}$ of $2c$ vertices and edges of $G_k$ where each $c\in \phi(x_c)\cap \phi(y_c)$ for each $c\in\{1,\ldots,C\}$. Ultimately, we want to use these to find a path in $H$ that contains at least two vertices of each colour.  We start by first establishing some properties of $R\subseteq V(G_k)$.

  Let $R_1:=z\in\bigcup_{c\in S}\{x_c,y_c\}$ and $R_2:=z\in\bigcup_{c\in \overline{S}}\{x_c,y_c\}$.

  \begin{clm}\label{far_from_home}
    For each $z\in R_1$ and $z'\in R_2$ with $s(z')\neq z$, $\dist_{G_k}(z,z')\ge \beta /2$.
  \end{clm}

  \begin{proof}
    Since $s(z')\neq z$, there exists some other $z''\in R_1\setminus\{z\}$ with $s(z')=s(z'')$.  Since $z,z''\in R_1$ each of $z$ and $z''$ were selected in Stage~1.  The rules for selecting vertices in Stage~1 guarantee that $\dist_{G_k}(z,z'')\ge \beta$.  By the triangle inequality $\dist(z, z')+\dist(z',z'') \ge \dist(z,z'')\ge \beta$.  Since $s(z')=z''$, $\dist_{G_k}(z', z'') \le \dist_{G_k}(z, z')$, so $\dist_{G_k}(z,z') \ge \dist(z',z'')/2 \ge \beta/2$.
  \end{proof}

@@

  \begin{proof}
    To prove this, we have to define $s(x)$ a little more carefully.
  \end{proof}

  \begin{clm}
    Any $5$-element subset of $R$ contains a pair $x,y$ with $\dist_{G_k}(x,y)\ge\beta/2$.
  \end{clm}

  \begin{proof}
    Let $R'$ be some $5$-element subset of $R$. $R'$ contains two or more elements of $R_1$ then the distance between these two elements is at most $\beta>\beta/2$.  Therefore, we may assume that $R'$ contains at most one element in $R_1$.  If $R'$ contains exactly one element $x$ in $R_1$ then, by \cref{far_from_home}, there is at most one other element in $R'$ whose distance to $x$ is at most $\beta/2$, and the three remaining elements have greater distance.  Therefore, $R'$ contains no elements in $R_1$.  Therefore, $S:=\{s(x):x\in R'\}$ is a five-element set whose pairwise distances are all at most $\beta$.  But this implies that two elements in $R'$ have distance less than $\beta$.  This is a contradiction because all points in $R'$ are in $R_1$ and all pairs of points in $R_1$ have distance at least $\beta$.\todo{Something is off by a factor of 2.}
  \end{proof}

  \begin{cor}
      In any $\beta/2\times \beta/2$ subgrid $G_k$ contains at most $4$ elements of $R$.
  \end{cor}

@@

    \begin{lem}
        $G_k$ contain a simple path $P$ that contains every edge and vertex in $R$.  Furthermore, $P$ bends at each bent vertex in $R$ and $P$ goes straight at each straight vertex in $R$. 
    \end{lem}
\end{proof}

@@
